---
title: "Introduction to raddr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to raddr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(raddr)
```

raddr reports what an IP address literal means under each of the standards and
implementations that disagree about it. The parsers that do that are not written
yet; what exists today is the value type they will return.

## The address type

`raddr_address()` is a low-level constructor over four 32-bit words. IPv4 lives
in the last word:

```{r}
localhost <- raddr_address(0L, 0L, 0L, 2130706433L, "v4")
localhost
```

The family has three states, not two, so the 4-in-6 form is not confused with
the IPv4 address it carries:

```{r}
mapped <- raddr_address(0L, 0L, 65535L, 2130706433L, "v6_4in6")
mapped == localhost
```

## The zone travels beside the bits

An RFC 4007 zone ID names an interface, not an address, so it is stored
separately and does not participate in equality. Two addresses that differ only
by zone are the same address:

```{r}
lo0 <- raddr_address(-25165824L, 0L, 0L, 1L, "v6", zone = "lo0")
en0 <- raddr_address(-25165824L, 0L, 0L, 1L, "v6", zone = "en0")

lo0 == en0
addr_zone(lo0)
```

## Ordering is total

IPv4 sorts before IPv6, so `sort()` works on a vector that mixes families:

```{r}
sort(c(mapped, localhost, lo0))
```

Formatting is deliberately unfinished: the output above is the fully expanded
form, not the RFC 5952 canonical form.
